\LARGE\textbf{Designing Studies and Evaluating Research Results:\\ Type M and Type S Errors for\\ Pearson Correlation Coefficient }

### 2021-02-12 Documents/Paper_main/Paper_main.Rnw
end review lakens

\begin{abstract}
It is widely appreciated that many studies in psychological science suffer from low statistical power. One of the consequences of analyzing underpowered studies with thresholds of statistical significance, is a high risk of finding exaggerated effect size estimates, in the right or in the wrong direction. These inferential risks can be directly quantified in terms of Type M (magnitude) error and Type S (sign) error, which directly communicate the consequences of design choices on effect size estimation. Given a study design, Type M error is the factor by which a statistically significant effect is on average exaggerated. Type S error is the probability to find a statistically significant result in the opposite direction to the plausible one. Ideally, these errors should be considered during a \emph{prospective design analysis} in the design phase of a study to determine the appropriate sample size. However, they can also be considered when evaluating studies’ results in a \emph{retrospective design analysis}. In the present contribution we aim to facilitate the considerations of these errors in the research practice in psychology. For this reason we illustrate how to consider Type M and Type S errors in a design analysis using one of the most common effect size measures in psychology: Pearson correlation coefficient. We provide various examples and make the R functions freely available to enable researchers to perform design analysis for their research projects.
\end{abstract} \hspace{10pt}

@@

Psychological science is increasingly committed to scrutinize its published findings by promoting large-scale replication efforts, where the protocol of a previous study is repeated as closely as possible with a new sample\parencite{camererEvaluatingReplicabilityLaboratory2016, camererEvaluatingReplicabilitySocial2018, ebersoleManyLabsEvaluating2016, kleinInvestigatingVariationReplicability2014, kleinManyLabsInvestigating2018, opensciencecollaborationEstimatingReproducibilityPsychological2015}. Interestingly, many replication studies found smaller effects than originals \parencite{camererEvaluatingReplicabilitySocial2018, opensciencecollaborationEstimatingReproducibilityPsychological2015} and among many possible explanations, one relates to a feature of study design: statistical power. In particular, it is plausible for original studies to have lower statistical power than their replications. In case of underpowered studies, we are usually aware of the lower probability of detecting an effect if this actually exists, but the less obvious consequences on effect size estimation are often neglected. When underpowered studies are analyzed using thresholds, such as statistical significance levels, effects passing such thresholds have to exaggerate the true effect size \parencite{buttonPowerFailureWhy2013, gelmanTypeErrorMight2017, ioannidisWhyMostDiscovered2008, ioannidisEmergenceLargeTreatment2013, laneEstimatingEffectSize1978}. Indeed, as will be extensively shown below, in underpowered studies only large effects correspond to values that can reject the null hypothesis and be statistically significant. As a consequence, if the original study was underpowered and found an exaggerated estimate of the effect, the replication effect will likely be smaller.

The concept of statistical power finds its natural development in the Neyman-Pearson framework of statistical inference and this is the framework that we adopt in this contribution. Contrary to the Null Hypothesis Significance Testing (NHST), the Neyman-Pearson approach requires to define both the \textit{Null Hypothesis} (i.e., usually, but not necessarily, the absence of an effect) and the \textit{Alternative Hypothesis} (i.e., the magnitude of the expected effect). Further discussion on the Neyman and Pearson approach and a comparison with the NHST is available in \textcite{altoeEnhancingStatisticalInference2020} and \textcite{gigerenzerNullRitualWhat2004}.
When conducting hypothesis testing, we usually consider two inferential risks:the Type I error (i.e., the probability $\alpha$ of rejecting the Null Hypothesis if this is true) and the Type II error (i.e., the probability $\beta$ of not rejecting the Null Hypothesis if this is false). Then, statistical power is defined as the probability 1-$\beta$ of finding a statistically significant result if the Alternative Hypothesis is true. All this leads to a narrow focus on statistical significance in hypothesis testing, overlooking another important aspect of statistical inference, namely, the effect size estimation.

When effect size estimation is conditioned on the statistical significance (i.e., effect estimates are evaluated only if their p-values are lower than $\alpha$), effect size exaggeration is a corollary consequence of low statistical power that might not be evident at first.This point can be highlighted considering the Type M (magnitude) and Type S (sign) errors characterizing a study design \parencite{gelmanPowerCalculationsAssessing2014}. Given a study design (i.e., sample size, statistical test directionality, $\alpha$ level and plausible effect size formalization), Type M error indicates the factor by which a statistically significant effect would be, on average, exaggerated. Type S error indicates the probability to find a statistically significant effect in the opposite direction to the one considered plausible. The analysis that researchers perform to evaluate the Type M and Type S errors in their research practice is called \emph{design analysis}, given the special focus posed into considering the design of a study \parencite{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014}.

Both errors are defined starting from a reasoned guess on the plausible magnitude and direction of the effect under study, which is called \emph{plausible effect size} \parencite{gelmanPowerCalculationsAssessing2014}. A plausible effect size is an assumption the researchers make about which is the expected effect in the population. This should not be based on some noisy results from a pilot study but, rather, it could derive from an extensive evaluation of the literature (e.g., theoretical or literature reviews and meta-analyses).  When considering the published literature to define the plausible effect size, however, it is important to take into account the presence of publication bias \parencite{francoPublicationBiasSocial2014} and consider techniques for adjusting for the possible inflation of effect size estimates \parencite{andersonSampleSizePlanningMore2017}. For example if, after taking into account possible inflations, all the main results in a given topic, considering a specific experimental design indicate that the correlation ranges between $r = .15$ and $r =.25$, we could reasonably choose as plausible effect size a value within this range. Or even better, we could consider multiple values to evaluate the results in different scenarios. Note that the definition of a plausible effect size is inevitably highly context dependent so any attempt to provide reference values would not be useful , instead it would prevent researchers from reasoning about the phenomenon of interest. Even in extreme cases where no previous information is available, which would question the exploratory/confirmatory nature of the study, researchers could still evaluate which effect would be considered relevant (e.g., from clinical or economic perspective) and define the plausible effect size according to that.

Why do these errors matter? The concepts of Type M and Type S errors allow enhancing researchers’ awareness on a complex process such as statistical inference. It gives a further tool that can be used to interpret and understand the important role of statistical power both when designing a new study or when evaluating previous results from the literature. In particular, it highlights the unwanted (and often overlooked) consequences on effect estimation when filtering for statistical significance in underpowered studies. The exaggeration of effect sizes, in the right or in the wrong direction, has important implications on a theoretical and applied level. On a theoretical level, studies’ designs with high Type M and Type S errors can foster distorted expectations on the effect under study, triggering a vicious cycle for the planning of future studies. This point is relevant also for the design of replication studies, which could turn out to be underpowered if they do not take into account possible inflations of the original effect \parencite{buttonPowerFailureWhy2013}. When studies are used to inform policymaking and real-world interventions, implications can go beyond the academic research community and can impact society at large. In these settings, we could assist to a “hype and disappointment cycle” \parencite{gelmanOverconfidenceResearchCertainty2019}, where true effects turn out to be much less impressive than expected. This can produce undesirable consequences on people’s lives, a consideration that invites researchers to assume responsibility in effectively communicating the risks related to effects quantification.

Type M (magnitude) and Type S (sign) errors are not widely known in the psychological research community but their consideration during the research process has the potential to improve current research practices, for example, by increasing the awareness that design choices have on possible studies' results. In a previous work we illustrated Type M and Type S errors using Cohens’\emph{d} as a measure of effect size \parencite{altoeEnhancingStatisticalInference2020}. The purpose of the present contribution is to further increase familiarity with Type M and Type S errors, considering another common effect size measures in psychology: Pearson correlation coefficient, $\rho$.  Weaim to provide an accessible introduction to the Design Analysis framework and enhance Type M and Type S errors understanding by means of several examples. The rest of this article is organized as follows: introduction to Type M and Type S errors; description of what is a design analysis and how to conduct one; analysis of Type S and Type M errors when varying alpha levels and hypothesis directionality. Moreover, the two appendices present further implications of design analysis for Pearson correlation (Appendix \ref{sec:appendix_a}) and an extensive illustration of the R functions for design analysis for Pearson correlation (Appendix \ref{sec:appendix_b}).

@@

Pearson correlation coefficient is a standardized effect size measure indicating the strength and the direction of the relationship between two continuous variables \parencite{cohenStatisticalPowerAnalysis1988, ellisEssentialGuideEffect2010}. Even though the correlation coefficient is widely known, we briefly go over its main features using an example. Imagine that we were interested to measure the relationship between anxiety and depression in a population and we plan a study with \emph{n} participants, where, for each participant, we measure the level of anxiety (i.e., variable X) and the level of depression (i.e., variable Y). At the end of the study we will have n pairs of values X and Y. The correlation coefficient helps us answer the questions: how strong is the linear relationship between anxiety and depression in this population? Is the relationship positive or negative? Correlation ranges from -1 to +1, indicating respectively two extreme scenarios of perfect negative relationship and perfect positive relationship\footnote{Correlation indicates a relationship between variables but does not imply causation. We do not discuss
this relevant aspect here but we refer the interested reader to \parencite{rohrerThinkingClearlyCorrelations2018}}. Since the correlation coefficient is a dimensionless number, it is a signal to noise ratio where the signal is given by the covariance between the two variables ($cov(x,y)$) and the noise is expressed by the product between the standard deviations of the two variables ($S_x  S_y$; see Formula\ref{eq:correlation}). Following the conventional standards, in this contribution we will use the symbol $\rho$ to indicate the correlation in the population and the symbol \emph{r} to indicate the value measured in a sample.
\begin{equation}
\label{eq:correlation}
r=\frac{cov(x,y)}{S_x\ S_y}.
\end{equation}

Magnitude and sign are two important features characterizing Pearson correlation coefficient and effect size measures in general. And, when estimating effect sizes, errors could be committed exactly regarding these two aspects. \textcite{gelmanPowerCalculationsAssessing2014} introduced two indexes to quantify these risks:
\begin{itemize}
  \item{Type M error, where M stands for magnitude, is also called Exaggeration Ratio - the factor by which a statistically significant effect is on average exaggerated.}
  \item{Type S error, sign - the probability to find a statistically significant result in the opposite direction to the plausible one.}
\end{itemize}

How are these errors computed? In the next paragraphs we approach this question preferring an intuitive perspective. For a formal definition of these errors, we refer the reader to \textcite{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014, luNoteTypeErrors2018}. Take as an example the previous fictitious study on the relationship between anxiety and depression and imagine we decide to sample 50 individuals (sample size, n = 50) and to set the $\alpha$ level to 5\% and to perform a two-tailed test. On the basis of theoretical considerations, we expect the plausibly true correlation in the population to be quite strong and positive which we formalize as $\rho$ = .50. In order to evaluate the Type M and Type S errors in this research design, imagine repeating the same study many times with new samples drawn from the same population and, for each study, register the observed correlation (r) and the corresponding p-value.

@@

The central and possibly most difficult point in this process is reasoning on what could be the plausible magnitude and direction of the effect of interest. This critical process, which is central also in traditional power analysis, is an opportunity for researchers to aggregate, formalize and incorporate prior information on the phenomenon under investigation \parencite{gelmanPowerCalculationsAssessing2014}. What is plausible can be determined on theoretical grounds, using expert knowledge elicitation techniques and consulting literature reviews and meta-analysis, always taking into account the presence of effect sizes inflation in the published literature \parencite{andersonBestOftForgotten2019}. The plausible effect size approximates the true effect, which is never known but can be thought as “that which would be observed in a hypothetical infinitely large sample” \parencite[p. 642]{gelmanPowerCalculationsAssessing2014}. For a more exhaustive description on plausible effect size, we refer the interested reader to \textcite{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014}.

@@

Researchers can consider Type M and Type S errors in their practice by performing a \emph{design analysis} \parencite{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014}. Ideally, a design analysis should be performed when designing a study. In this phase it is specifically called \emph{prospective design analysis} and it can be used as a sample size planning strategy where statistical power is considered together with Type M and Type S errors. However, design analysis can also be beneficial to evaluate the inferential risks in studies that have already been conducted and where the study design is known. In these cases, Type M and Type S errors can support results interpretation by communicating the inferential risks in that research design. When design analysis happens at this later stage, it takes the name of \emph{retrospective design analysis}. Note that retrospective design analysis should not be confused with post-hoc power analysis. A retrospective design analysis defines the plausible effect size according to previous results in the literature or other information external to the study, whereas the post-hoc power analysis defines the plausible effect size based on the observed results in the study and it is a widely-deprecated practice \parencite{goodmanUsePredictedConfidence1994a, gelmanDonCalculatePosthoc2019a}.

In the following sections we illustrate how to perform prospective and retrospective design analysis using some examples. We developed two R functions\footnote{An R-package was subsequently developed and now is available on CRAN, PRDA: Conduct a Prospective or Retrospective Design Analysis \url{https://cran.r-project.org/web/packages/PRDA/index.html}.} to perform design analysis for Pearson correlation, which are available at the page \url{https://osf.io/9q5fr/}. The function to perform a prospective design analysis is \texttt{pro\_r()}. It requires as input the plausible effect size (\texttt{rho}), the statistical power (\texttt{power}), the directionality of the test (\texttt{alternative}) which can be set as: “\texttt{two.sided}”, “\texttt{less}” or “\texttt{greate}r” . Type I error rate (\texttt{sig\_level}) is set as default at 5\% and can be changed by the user. The \texttt{pro\_r()} function returns the necessary sample size to achieve the desired statistical power, Type M error rate, the Type S error probability, and the critical value(s) above which a statistically significant result can be found. The function to perform retrospective design analysis is \texttt{retro\_r()}. It requires as input the plausible effect size, the sample size used in the study, and the directionality of the test that was performed. Also in this case, Type I error rate is set as default at 5\% and can be changed by the user.  The function \texttt{retro\_r()} returns the Type M error rate, the Type S error probability, and the critical value(s)\footnote{\emph{Critical value} is the name usually employed in hypotheses testing within the Neyman-Pearson framework. In the research practice this is also known as the \emph{Minimal Statistically Detectable Effect} \parencite{cookAssessingMethodsSpecify2014, phillipsStatisticalSignificanceSediment2001}} . For further details regarding the R functions refer to Appendix~\ref{sec:appendix_b}.

@@

We consider the study published in \emph{Science} by \textcite{eisenbergerDoesRejectionHurt2003} entitled: “Does Rejection Hurt? An fMRI Study of Social Exclusion”. The research question originated from the observation that the Anterior Cingulate Cortex (ACC) is a region of the brain known to be involved in the experience of physical pain. Could pain from social stimuli, such as social exclusion, share similar neural underpinnings? To test this hypothesis, 13 participants were recruited and each one had to play a virtual game with other two players while undergoing functional Magnetic Resonance Imaging (fMRI). The other two players were fictitious, and participants were actually playing against a computer program. Players had to toss a virtual ball among each other in three conditions: social inclusion, explicit social exclusion and implicit social exclusion. In the social inclusion condition the participant regularly received the ball. In the explicit social exclusion condition the participant was told that, due to technical problems, he was not going to play that round. In the implicit social exclusion condition, the participant experienced being intentionally left out from the game by the other two players. At the end of the experiment, each participant completed a self-report measure regarding their perceived distress when they were intentionally left out by the other players. Considering only the implicit social exclusion condition, a correlation coefficient was estimated between the measure of distress and neural activity in the Anterior Cingulate Cortex. As suggested by the large and statistically significant correlation coefficient between perceived distress and activity in the ACC, r = .88, p < .005 \parencite[p. 291]{eisenbergerDoesRejectionHurt2003}, authors concluded that social and physical pain seem to share similar neural underpinnings.

Before proceeding to the retrospective design analysis, we refer the interested reader to some background history regarding this study. This was one of the many studies included in the famous paper “Puzzlingly High Correlations in fMRI Studies of Emotion, Personality, and Social Cognition” \parencite{vulPuzzlinglyHighCorrelations2009} which raised important issues regarding the analysis of neuroscientific data. In particular, this paper noted that the magnitude of correlation coefficients between fMRI measures and behavioral measures were beyond what could be considered plausible. We refer the interested reader also to the commentary by \textcite{yarkoniBigCorrelationsLittle2009}, who noted that the implausibly high correlations in fMRI studies could be largely explained by the low statistical power of experiments.

A retrospective design analysis should start with thorough reasoning on the plausible size and direction of the effect under study. In order to produce valid inferences, a lot of attention should be devoted to this point by integrating external information. For the sake of this example, we turn to the considerations made by \textcite{vulSuspiciouslyHighCorrelations2017} who suggested correlations between personality measures and neural activity to be likely around $\rho=.25$.  A correlation of $\rho=.50$ was deemed plausible but optimistic and a correlation of $\rho=.75$ was considered theoretically plausible but unrealistic.

@@

To perform a retrospective design analysis on the case study, we need information on the research design and the plausible effect size. Based on the previous considerations, we set the plausible effect size to be $\rho=.25$. Information on the sample size was not available in the original study \parencite{eisenbergerDoesRejectionHurt2003} and was retrieved from \textcite{vulPuzzlinglyHighCorrelations2009} to be $n=13$. The $\alpha$ level and the directionality of the test were not reported in the original study, so for the purpose of this example we will consider $\alpha=.05$  and a two-tailed test. Given this study design, what are the inferential risks in terms of effect size estimation?

>>>>

\begin{abstract}
It is widely appreciated that many studies in psychological science suffer from low statistical power. One of the consequences of analyzing underpowered studies with thresholds of statistical significance is a high risk of finding exaggerated effect size estimates, in the right or the wrong direction. These inferential risks can be directly quantified in terms of Type M (magnitude) error and Type S (sign) error, which directly communicate the consequences of design choices on effect size estimation. Given a study design, Type M error is the factor by which a statistically significant effect is on average exaggerated. Type S error is the probability to find a statistically significant result in the opposite direction to the plausible one. Ideally, these errors should be considered during a \emph{prospective design analysis} in the design phase of a study to determine the appropriate sample size. However, they can also be considered when evaluating studies’ results in a \emph{retrospective design analysis}. In the present contribution, we aim to facilitate the considerations of these errors in the research practice in psychology. For this reason, we illustrate how to consider Type M and Type S errors in a design analysis using one of the most common effect size measures in psychology: Pearson correlation coefficient. We provide various examples and make the R functions freely available to enable researchers to perform design analysis for their research projects.
\end{abstract} \hspace{10pt}

@@

Psychological science is increasingly committed to scrutinizing its published findings by promoting large-scale replication efforts, where the protocol of a previous study is repeated as closely as possible with a new sample\parencite{camererEvaluatingReplicabilityLaboratory2016, camererEvaluatingReplicabilitySocial2018, ebersoleManyLabsEvaluating2016, kleinInvestigatingVariationReplicability2014, kleinManyLabsInvestigating2018, opensciencecollaborationEstimatingReproducibilityPsychological2015}. Interestingly, many replication studies found smaller effects than originals \parencite{camererEvaluatingReplicabilitySocial2018, opensciencecollaborationEstimatingReproducibilityPsychological2015} and among many possible explanations, one relates to a feature of study design: statistical power. In particular, it is plausible for original studies to have lower statistical power than their replications. In the case of underpowered studies, we are usually aware of the lower probability of detecting an effect if this exists, but the less obvious consequences on effect size estimation are often neglected. When underpowered studies are analyzed using thresholds, such as statistical significance levels, effects passing such thresholds have to exaggerate the true effect size \parencite{buttonPowerFailureWhy2013, gelmanTypeErrorMight2017, ioannidisWhyMostDiscovered2008, ioannidisEmergenceLargeTreatment2013, laneEstimatingEffectSize1978}. Indeed, as will be extensively shown below, in underpowered studies only large effects correspond to values that can reject the null hypothesis and be statistically significant. As a consequence, if the original study was underpowered and found an exaggerated estimate of the effect, the replication effect will likely be smaller.

The concept of statistical power finds its natural development in the Neyman-Pearson framework of statistical inference and this is the framework that we adopt in this contribution. Contrary to the Null Hypothesis Significance Testing (NHST), the Neyman-Pearson approach requires to define both the \textit{Null Hypothesis} (i.e., usually, but not necessarily, the absence of an effect) and the \textit{Alternative Hypothesis} (i.e., the magnitude of the expected effect). Further discussion on the Neyman and Pearson approach and a comparison with the NHST is available in \textcite{altoeEnhancingStatisticalInference2020} and \textcite{gigerenzerNullRitualWhat2004}.
When conducting hypothesis testing, we usually consider two inferential risks: the Type I error (i.e., the probability $\alpha$ of rejecting the Null Hypothesis if this is true) and the Type II error (i.e., the probability $\beta$ of not rejecting the Null Hypothesis if this is false). Then, statistical power is defined as the probability 1-$\beta$ of finding a statistically significant result if the Alternative Hypothesis is true. All this leads to a narrow focus on statistical significance in hypothesis testing, overlooking another important aspect of statistical inference, namely, the effect size estimation.

When effect size estimation is conditioned on the statistical significance (i.e., effect estimates are evaluated only if their p-values are lower than $\alpha$), effect size exaggeration is a corollary consequence of low statistical power that might not be evident at first. This point can be highlighted considering the Type M (magnitude) and Type S (sign) errors characterizing a study design \parencite{gelmanPowerCalculationsAssessing2014}. Given a study design (i.e., sample size, statistical test directionality, $\alpha$ level and plausible effect size formalization), Type M error, also known as \textit{Exaggeration Ratio}, indicates the factor by which a statistically significant effect would be, on average, exaggerated. Type S error indicates the probability to find a statistically significant effect in the opposite direction to the one considered plausible. The analysis that researchers perform to evaluate the Type M and Type S errors in their research practice is called \emph{design analysis}, given the special focus posed into considering the design of a study \parencite{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014}.

Both errors are defined starting from a reasoned guess on the plausible magnitude and direction of the effect under study, which is called \emph{plausible effect size} \parencite{gelmanPowerCalculationsAssessing2014}. A plausible effect size is an assumption the researchers make about which is the expected effect in the population. This should not be based on some noisy results from a pilot study but, rather, it could derive from an extensive evaluation of the literature (e.g., theoretical or literature reviews and meta-analyses).  When considering the published literature to define the plausible effect size, however, it is important to take into account the presence of publication bias \parencite{francoPublicationBiasSocial2014} and consider techniques for adjusting for the possible inflation of effect size estimates \parencite{andersonSampleSizePlanningMore2017}. For example if, after taking into account possible inflations, all the main results in a given topic, considering a specific experimental design indicate that the correlation ranges between $r = .15$ and $r =.25$, we could reasonably choose as plausible effect size a value within this range. Or even better, we could consider multiple values to evaluate the results in different scenarios. Note that the definition of the plausible effect size is inevitably highly context-dependent so any attempt to provide reference values would not be useful, instead, it would prevent researchers from reasoning about the phenomenon of interest. Even in extreme cases where no previous information is available, which would question the exploratory/confirmatory nature of the study, researchers could still evaluate which effect would be considered relevant (e.g., from a clinical or economic perspective) and define the plausible effect size according to it.

Why do these errors matter? The concepts of Type M and Type S errors allow enhancing researchers’ awareness of a complex process such as statistical inference. It gives a further tool that can be used to interpret and understand the important role of statistical power both when designing a new study or when evaluating previous results from the literature. In particular, it highlights the unwanted (and often overlooked) consequences on effect estimation when filtering for statistical significance in underpowered studies. The exaggeration of effect sizes, in the right or the wrong direction, has important implications on a theoretical and applied level. On a theoretical level, studies’ designs with high Type M and Type S errors can foster distorted expectations on the effect under study, triggering a vicious cycle for the planning of future studies. This point is relevant also for the design of replication studies, where the possible inflation of the original effect should be taken into account together with the smallest effect size of interest to limit the chances of being underpowered \parencite{buttonPowerFailureWhy2013}. When studies are used to inform policymaking and real-world interventions, implications can go beyond the academic research community and can impact society at large. In these settings, we could assist to a “hype and disappointment cycle” \parencite{gelmanOverconfidenceResearchCertainty2019}, where true effects turn out to be much less impressive than expected. This can produce undesirable consequences on people’s lives, a consideration that invites researchers to assume responsibility in effectively communicating the risks related to effects quantification.

To our knowledge, Type M (magnitude) and Type S (sign) errors are not widely known in the psychological research community but their consideration during the research process has the potential to improve current research practices, for example, by increasing the awareness that design choices have on possible studies' results. In a previous work, we illustrated Type M and Type S errors using Cohen's’\emph{d} as a measure of effect size \parencite{altoeEnhancingStatisticalInference2020}. The purpose of the present contribution is to further increase the familiarity with Type M and Type S errors, considering another common effect size measures in psychology: Pearson correlation coefficient, $\rho$. We aim to provide an accessible introduction to the Design Analysis framework and enhance Type M and Type S errors understanding using several educational examples. The rest of this article is organized as follows: introduction to Type M and Type S errors; description of what is a design analysis and how to conduct one; analysis of Type S and Type M errors when varying alpha levels and hypothesis directionality. Moreover, the two appendices present further implications of design analysis for Pearson correlation (Appendix \ref{sec:appendix_a}) and an extensive illustration of the R functions for design analysis for Pearson correlation (Appendix \ref{sec:appendix_b}).

@@

Pearson correlation coefficient is a standardized effect size measure indicating the strength and the direction of the relationship between two continuous variables \parencite{cohenStatisticalPowerAnalysis1988, ellisEssentialGuideEffect2010}. Even though the correlation coefficient is widely known, we briefly go over its main features using an example. Imagine that we were interested to measure the relationship between anxiety and depression in a population and we plan a study with \emph{n} participants, where, for each participant, we measure the level of anxiety (i.e., variable X) and the level of depression (i.e., variable Y). At the end of the study, we will have n pairs of values X and Y. The correlation coefficient helps us answer the questions: how strong is the linear relationship between anxiety and depression in this population? Is the relationship positive or negative? Correlation ranges from -1 to +1, indicating respectively two extreme scenarios of perfect negative relationship and perfect positive relationship\footnote{Correlation indicates a relationship between variables but does not imply causation. We do not discuss
this relevant aspect here but we refer the interested reader to \parencite{rohrerThinkingClearlyCorrelations2018}}. Since the correlation coefficient is a dimensionless number, it is a signal to noise ratio where the signal is given by the covariance between the two variables ($cov(x,y)$) and the noise is expressed by the product between the standard deviations of the two variables ($S_x  S_y$; see Formula\ref{eq:correlation}). In this contribution, following the conventional standards, we will use the symbol $\rho$ to indicate the correlation in the population and the symbol \emph{r} to indicate the value measured in a sample.
\begin{equation}
\label{eq:correlation}
r=\frac{cov(x,y)}{S_x\ S_y}.
\end{equation}

Magnitude and sign are two important features characterizing Pearson correlation coefficient and effect size measures in general. And, when estimating effect sizes, errors could be committed exactly regarding these two aspects. \textcite{gelmanPowerCalculationsAssessing2014} introduced two indexes to quantify these risks:
\begin{itemize}
  \item{Type M error, where M stands for magnitude, is also called Exaggeration Ratio - the factor by which a statistically significant effect is on average exaggerated.}
  \item{Type S error, sign - the probability to find a statistically significant result in the opposite direction to the plausible one.}
\end{itemize}
Note that, differently from the other inferential errors, Type M error is not a probability but rather a ratio indicating the average percentage of inflation.

How are these errors computed? In the next paragraphs, we approach this question preferring an intuitive perspective. For a formal definition of these errors, we refer the reader to \textcite{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014, luNoteTypeErrors2018}. Take as an example the previous fictitious study on the relationship between anxiety and depression and imagine we decide to sample 50 individuals (sample size, n = 50) and to set the $\alpha$ level to 5\% and to perform a two-tailed test. Based on theoretical considerations, we expect the plausibly true correlation in the population to be quite strong and positive which we formalize as $\rho$ = .50. To evaluate the Type M and Type S errors in this research design, imagine repeating the same study many times with new samples drawn from the same population and, for each study, register the observed correlation (r) and the corresponding p-value.

@@

The central and possibly most difficult point in this process is reasoning on what could be the plausible magnitude and direction of the effect of interest. This critical process, which is central also in traditional power analysis, is an opportunity for researchers to aggregate, formalize and incorporate prior information on the phenomenon under investigation \parencite{gelmanPowerCalculationsAssessing2014}. What is plausible can be determined on theoretical grounds, using expert knowledge elicitation techniques and consulting literature reviews and meta-analysis, always taking into account the presence of effect sizes inflation in the published literature \parencite{andersonBestOftForgotten2019}. The plausible effect size approximates the true effect, which is never known but can be thought of as “that which would be observed in a hypothetical infinitely large sample” \parencite[p. 642]{gelmanPowerCalculationsAssessing2014}. For a more exhaustive description of plausible effect size, we refer the interested reader to \textcite{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014}.

@@

Researchers can consider Type M and Type S errors in their practice by performing a \emph{design analysis} \parencite{altoeEnhancingStatisticalInference2020, gelmanPowerCalculationsAssessing2014}. Ideally, a design analysis should be performed when designing a study. In this phase, it is specifically called \emph{prospective design analysis} and it can be used as a sample size planning strategy where statistical power is considered together with Type M and Type S errors. However, design analysis can also be beneficial to evaluate the inferential risks in studies that have already been conducted and where the study design is known. In these cases, Type M and Type S errors can support results interpretation by communicating the inferential risks in that research design. When design analysis happens at this later stage, it takes the name of \emph{retrospective design analysis}. Note that retrospective design analysis should not be confused with post-hoc power analysis. A retrospective design analysis defines the plausible effect size according to previous results in the literature or other information external to the study, whereas the post-hoc power analysis defines the plausible effect size based on the observed results in the study and it is a widely-deprecated practice \parencite{goodmanUsePredictedConfidence1994a, gelmanDonCalculatePosthoc2019a}.

In the following sections, we illustrate how to perform prospective and retrospective design analysis using some examples. We developed two R functions\footnote{An R-package was subsequently developed and now is available on CRAN, PRDA: Conduct a Prospective or Retrospective Design Analysis \url{https://cran.r-project.org/web/packages/PRDA/index.html}.} to perform design analysis for Pearson correlation, which are available at the page \url{https://osf.io/9q5fr/}. The function to perform a prospective design analysis is \texttt{pro\_r()}. It requires as input the plausible effect size (\texttt{rho}), the statistical power (\texttt{power}), the directionality of the test (\texttt{alternative}) which can be set as: “\texttt{two.sided}”, “\texttt{less}” or “\texttt{greater}”. Type I error rate (\texttt{sig\_level}) is set as default at 5\% and can be changed by the user. The \texttt{pro\_r()} function returns the necessary sample size to achieve the desired statistical power, Type M error rate, the Type S error probability, and the critical value(s) above which a statistically significant result can be found. The function to perform retrospective design analysis is \texttt{retro\_r()}. It requires as input the plausible effect size, the sample size used in the study, and the directionality of the test that was performed. Also in this case, Type I error rate is set as default at 5\% and can be changed by the user.  The function \texttt{retro\_r()} returns the Type M error rate, the Type S error probability, and the critical value(s)\footnote{\emph{Critical value} is the name usually employed in hypotheses testing within the Neyman-Pearson framework. In the research practice, this is also known as the \emph{Minimal Statistically Detectable Effect} \parencite{cookAssessingMethodsSpecify2014, phillipsStatisticalSignificanceSediment2001}}. For further details regarding the R functions refer to Appendix~\ref{sec:appendix_b}.

@@

We consider the study published in \emph{Science} by \textcite{eisenbergerDoesRejectionHurt2003} entitled: “Does Rejection Hurt? An fMRI Study of Social Exclusion”. The research question originated from the observation that the Anterior Cingulate Cortex (ACC) is a region of the brain known to be involved in the experience of physical pain. Could pain from social stimuli, such as social exclusion, share similar neural underpinnings? To test this hypothesis, 13 participants were recruited and each one had to play a virtual game with other two players while undergoing functional Magnetic Resonance Imaging (fMRI). The other two players were fictitious, and participants were actually playing against a computer program. Players had to toss a virtual ball among each other in three conditions: social inclusion, explicit social exclusion and implicit social exclusion. In the social inclusion condition, the participant regularly received the ball. In the explicit social exclusion condition the participant was told that, due to technical problems, he was not going to play that round. In the implicit social exclusion condition, the participant experienced being intentionally left out from the game by the other two players. At the end of the experiment, each participant completed a self-report measure regarding their perceived distress when they were intentionally left out by the other players. Considering only the implicit social exclusion condition, a correlation coefficient was estimated between the measure of distress and neural activity in the Anterior Cingulate Cortex. As suggested by the large and statistically significant correlation coefficient between perceived distress and activity in the ACC, r = .88, p < .005 \parencite[p. 291]{eisenbergerDoesRejectionHurt2003}, authors concluded that social and physical pain seem to share similar neural underpinnings.

Before proceeding to the retrospective design analysis, we refer the interested reader to some background history regarding this study. This was one of the many studies included in the famous paper “Puzzlingly High Correlations in fMRI Studies of Emotion, Personality, and Social Cognition” \parencite{vulPuzzlinglyHighCorrelations2009} which raised important issues regarding the analysis of neuroscientific data. In particular, this paper noted that the magnitude of correlation coefficients between fMRI measures and behavioural measures were beyond what could be considered plausible. We refer the interested reader also to the commentary by \textcite{yarkoniBigCorrelationsLittle2009}, who noted that the implausibly high correlations in fMRI studies could be largely explained by the low statistical power of experiments.

A retrospective design analysis should start with thorough reasoning on the plausible size and direction of the effect under study. To produce valid inferences, a lot of attention should be devoted to this point by integrating external information. For the sake of this example, we turn to the considerations made by \textcite{vulSuspiciouslyHighCorrelations2017} who suggested correlations between personality measures and neural activity to be likely around $\rho=.25$.  A correlation of $\rho=.50$ was deemed plausible but optimistic and a correlation of $\rho=.75$ was considered theoretically plausible but unrealistic.

@@

To perform a retrospective design analysis on the case study, we need information on the research design and the plausible effect size. Based on the previous considerations, we set the plausible effect size to be $\rho=.25$. Information on the sample size was not available in the original study \parencite{eisenbergerDoesRejectionHurt2003} and was retrieved from \textcite{vulPuzzlinglyHighCorrelations2009} to be $n=13$. The $\alpha$ level and the directionality of the test were not reported in the original study, so for the purpose of this example, we will consider $\alpha=.05$  and a two-tailed test. Given this study design, what are the inferential risks in terms of effect size estimation?

We can use the R function \texttt{retro\_r()}, whose inputs and outputs are displayed in Figure~\ref{fig:retro_r}. In this study, the statistical power is .13, that is to say, there is a 13\% probability to reject the null hypothesis, if an effect of at least $\rho=|.25|$ exists. Consider this point together with the results obtained in the experiment: $r = .88$, $p < .005$ \parencite[p. 291]{eisenbergerDoesRejectionHurt2003}. It is clear that, even though the probability to reject the null hypothesis is low (power of 13\%), this event could happen. And when it does happen, it is tempting to believe that results are even more remarkable \parencite{gelmanStatisticalCrisisScience2014}. However, this design comes with serious inferential risks for the estimation of effect sizes, which could be grasped by presenting Type M and Type S errors. A glance at their value communicates that it is not impossible to find a statistically significant result, but when it does happen, the effect sizes could be largely overestimated - Type M = 2.58 - and maybe even in the wrong direction - Type S = .03. The Type M error rate of  2.58 indicates that a statistically significant correlation is on average about two and a half times the plausible value. In other words, statistically significant results emerging in such a research design will on average overestimate the plausible correlation coefficient by 160\%. The Type S error of  .03 suggests that there is a three percent probability to find a statistically significant result in the opposite direction, in this example, a negative relationship.
